Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy

### 2020-08-09 chapters_altFormat/ch_intro.tex
Added Betty's edits

Mankind has been continuously interested in the nature of the matter which composes the visible universe, and what are its constituents or building blocks. The seminal discovery was made in 1911 by Ernest Rutherford who scattered alpha particles from a gold foil using particles emitted by an alpha source from Marie Curie. His studies led him to conclude that atoms consist of a dense, positively charged atomic nucleus, surrounded by a dilute electron cloud. Later,  James Chadwick  discovered that the nucleus contained neutral particles -- neutrons  -- in addition to the known positively charged particle -- protons -- as the basic constituents of nuclear matter. Remarkably, the nucleus accounts for 99.9\% of the mass of the atom, and only \num{e-12} of the total volume, making it the densest stable form of matter on Earth.

Without an attractive force between nucleons, the Coulomb force between protons would render the nucleus unstable. It is the strong nuclear force which exerts a large attractive force  between nucleons over a very short range. This strong force is a manifestation of the fundamental interactions between the quarks which constitute each nucleon. Three quarks, of two types or \emph{flavors} labeled \emph{up} and \emph{down}, with nearly equal masses form each nucleon. At low energies the internal quark structure of nucleons is less important and nucleons can be thought of as fundamental particles. The strong force between nucleons is attractive only for a small distance of approximately \SIrange{1}{2}{\femto\metre} and becomes very repulsive at even shorter distances, making the nucleus very difficult to compress. It is for this reason that the distance between any pair of nucleons is a near constant value, and therefore the central density value is relatively constant over a wide range of nuclei, from the lightest to the heaviest nuclei \cite{krane}. This density is referred to as the  \emph{saturation density}, since the central density seems to saturate at a common value, namely, $\rho_0 = \SI{1.7e14}{\gram\per\centi\metre\cubed}$ or \num{0.16} nucleons \si{\per\femto\metre\cubed}.

@@

Since the strong force makes the inter-nucleon distance approximately constant, the volume the nucleus is related to the total number of nucleons and the volume, $A$. Therefore the first \emph{volume} term $a_V$ is proportional to $A$. Since the  constituent quarks that make of the nucleons are of nearly equal mass, there is a fundamental symmetry of the strong interaction which the strong interaction between any pair of nuclei is very similar, regardless of the particular flavor of quark composition. This symmetry is called \emph{isospin symmetry}. The term isospin was chosen because this symmetry was similar to the spin symmetry of electrons, which have spin up and down projections but are otherwise indistinguishable, unless in the presence of a symmetry breaking field.  The isospin mathematics are identical to the spin mathematics describing whether nucleons have an isospin of 1/2 and with projections of  +1/2 for protons and -1/2 for neutrons.

Continuing on with the Bethe-Weizacher formula, we see that there are several correction terms accounting for such effects like the surface, $a_s$, where nucleons near the surface have fewer surrounding neighbors as nucleons inside. Since the volume of the nuclei scales as $A$, radius of the nuclei would scale as $A^{1/3}$. Also there is a correction accounting for the Coulomb force, $a_c$, which is proportional to the radial extent of the nuclei $R^{-1}$; which scales as $A^{1/3}$. The asymmetry term, $a_A$, is related to the difference in energy between adding a neutron or a proton; it is typically referred to as the \emph{symmetry energy}. Although the strong force does not distinguish between protons and neutrons the overall wave function is made up of a combination of the spatial, spin, and isospin wave functions. Since nucleons are fermions,the total wave function must be anti-symmetric under particle exchange. Because of this, each spatial and spin state of a nucleon can only be occupied by one neutron and one proton, because there still is isospin.  Because of this it is more energetically favorable to form neutron-proton pairs than to become more and more neutron or proton rich. Thus the asymmetry term above is proportional to the factor $(N - Z)^2/A$ which comes directly from Pauli blocking. While Eq.~\ref{eq:semiEmp} describes normal nuclear matter, at saturation density, quite well, a more general description must be made when going to higher or lower densities.

@@

\begin{equation}
P = -\Big(\frac{\delta E}{\delta V}\Big)\vert_{T=0,N}
\label{eq:pressEos}
\end{equation}

@@

\section{Heavy Ion Collisions}
Besides observing neutron stars directly, heavy-ion collisions (HIC) provide the only way we can probe the density dependence of the symmetry energy in the laboratory setting. When two nuclei collide, in the very early stages they compress to form a high density region where the nuclei overlap. This momentary density can reach up to $3\rho_0$ depending on the incident beam energy, but only a small volume attains such densities.  HICs also provide the only way we can probe the isospin asymmetry dependence of the nuclear EoS. This is accomplished by using radioactive neutron-rich beams to collide on stable targets.

@@

\section{Pion Observable}
\label{sec:pionObs}
Another approach is to find an observable that can be detected with high efficiencies, unlike the neutron, and is more sensitive to just the high density region. For these reasons pions are a very interesting observable. Pions are produced though an intermediate step via production of $\Delta$ resonances. Here the reaction $ NN \leftrightarrow \Delta N$ takes place, where nucleon-nucleon collisions form $\Delta$(1232) baryon resonance from one of the nucleons, which then decays shortly thereafter into a pion $\Delta \leftrightarrow \pi N$. The threshold for  an on-shell $\Delta$ resonance production in nucleon-nucleon collisions, with a mass of \SI{1232}{\mega\electronvolt\per\clight\squared}, corresponds to a laboratory beam of \SI{290}{\MeVA} in a fixed target experiment. In large nuclei, the internal motion of the nucleons is substantial and this modifies the production rates. In the Fermi gas model, nucleons are arranged filling up higher energy levels up until they reach the Fermi energy. The associated Fermi velocity can add to the beam velocity and actually be above the threshold condition, even when the beam velocity is below the threshold \cite{fermiEnergy}.

@@

\section{Motivation}
In an effort to answer the high density behavior of the symmetry energy, we designed a new detector and a set of experiments of Sn + Sn collisions at \SI{270}{\MeVA}. Here we utilized inverse kinematics where the beam is made of a radio-active neutron rich beam, impinging on a stable Sn target. We can probe the isospin asymmetry of the symmetry energy by changing the neutron-proton asymmetry of the incoming beam. Pion production has been studied before in stable beams for beam energies of \SI{400}{\MeVA} and above \cite{fopi}. Yet in these previous experiments, only the total pion yields were published and no pion spectra were published. It was also known that the efficiency analysis of the FOPI collaboration which was particularly difficult especially for low energy pions. In fact the published pion multiplicities are not the measured values but rather extrapolated to low pion energies. We know now that below the cutoff cited in \cite{fopi}, approximately 35\% of the $\pi^+$ multiplicity and 55\% of the $\pi^-$ multiplicity, showing the importance of measuring down to low pion energies. While the pion production increases with the beam energy, this increase is countered by the fact that the effects of the symmetry energy decrease with energy \cite{fopi}. This is because available energy to produce pions is larger and therefore less sensitive to the relatively small symmetry energy.  The goal of this Dissertation was to design and build a high efficiency detector in order to measure pion and light charged particle spectra resulting from HICs. To do this a new Time Projection Chamber (TPC) was made called the SAMURAI pion Reconstruction Ion Tracker (\spirit) TPC.

>>>>

Mankind has been continuously interested in the nature of the matter which composes the visible universe, and what are its constituents or building blocks. The seminal discovery was made in 1911 by Ernest Rutherford who scattered Alpha particles from a gold foil using particles emitted by an Alpha source from Marie Curie. His studies led him to conclude that atoms consist of a dense, positively charged atomic nucleus, surrounded by a dilute electron cloud. Later,  James Chadwick  discovered that the nucleus contained neutral particles -- neutrons  -- in addition to the known positively charged particle -- protons -- as the basic constituents of nuclear matter. Remarkably, the nucleus accounts for 99.9\% of the mass of the atom, and only \num{e-12} of the total volume, making it the most dense stable form of matter on Earth.

Without an attractive force between nucleons, the Coulomb force between protons would render the nucleus unstable. It is the strong nuclear force which exerts a large attractive force  between nucleons over a very short range. This strong force is a manifestation of the fundamental interactions between the quarks which constitute each nucleon. Three quarks, of two types or \emph{flavors} labeled \emph{up} and \emph{down}, with nearly equal masses form each nucleon. At low energies the internal quark structure of nucleons is less important and nucleons can be thought of as fundamental particles. The strong force between nucleons is attractive only for a small distance of approximately \SIrange{1}{2}{\femto\metre} and becomes very repulsive at even shorter distances, making the nucleus very difficult to compress. It is for this reason that the distance between any pair of nucleons is nearly a constant value, and therefore the central density value is relatively constant over a wide range of nuclei, from the lightest to the heaviest nuclei \cite{krane}. This density is referred to as the  \emph{saturation density}, since the central density seems to saturate at a common value, namely, $\rho_0 = \SI{1.7e14}{\gram\per\centi\metre\cubed}$ or \num{0.16} nucleons \si{\per\femto\metre\cubed}.

@@

Since the strong force makes the inter-nucleon distance approximately constant, the volume term ,$a_V$, is proportional to the volume of the nucleus which is related to the total number of nucleons $A$. Since the  constituent quarks that make of the nucleons are of nearly equal mass, there is a fundamental symmetry of the strong interaction which the strong interaction between any pair of nuclei is very similar, regardless of the particular flavor of quark composition. This symmetry is called \emph{isospin symmetry}. The term isospin was chosen because this symmetry is similar to the spin symmetry of electrons, which have spin up and down projections but are otherwise indistinguishable, unless in the presence of a symmetry breaking field.  The isospin mathematics are identical to the spin mathematics describing whether nucleons have an isospin of 1/2 and with projections of  +1/2 for protons and -1/2 for neutrons.

Continuing on with the Bethe-Weizacher formula, we see that there are several correction terms accounting for such effects like the surface, $a_s$, where nucleons near the surface have fewer surrounding neighbors as nucleons inside. Since the volume of the nuclei scales as $A$, radius of the nuclei would scale as $A^{\frac{1}{3}}$. Also there is a correction accounting for the Coulomb force, $a_c$, which is proportional to the radial extent of the nuclei $R^{-1}$; which scales as $A^{\frac{1}{3}}$. The asymmetry term, $a_A$, is related to the difference in energy between adding a neutron or a proton; it is typically referred to as the \emph{symmetry energy}. Although the strong force does not distinguish between protons and neutrons the overall wave function is made up of a combination of the spatial, spin, and isospin wave functions. Since nucleons are fermions,the total wave function must be anti-symmetric under particle exchange. Because of this, each spatial and spin state of a nucleon can only be occupied by one neutron and one proton, because there still is isospin.  It is therefore more energetically favorable to form neutron-proton pairs than to become more and more neutron or proton rich. Thus the asymmetry term above is proportional to the factor $(N - Z)^2/A$ which comes directly from Pauli blocking. While Eq.~\ref{eq:semiEmp} describes normal nuclear matter, at saturation density, quite well, a more general description must be made when going to higher or lower densities.

@@

\begin{equation}
P = -\Big(\frac{\delta E}{\delta V}\Big)\vert_{T=0,N} ,
\label{eq:pressEos}
\end{equation}

@@

\section{Heavy Ion Collisions}
Besides observing neutron stars directly, heavy-ion collisions (HIC) provide the only way we can probe the density dependence of the symmetry energy in the laboratory setting. When two nuclei collide, in the very early stages they compress to form a high density region where the nuclei overlap. This momentary density can reach up to $3\rho_0$ ,depending on the incident beam energy, over a small volume where the nuclei overlap.  HICs also provide the only way we can probe the isospin asymmetry dependence of the nuclear EoS. This is accomplished by using radioactive neutron-rich beams to collide on stable targets.

@@

\section{Pion Observable}
\label{sec:pionObs}
Another approach is to find an observable that can be detected with high efficiencies, unlike the neutron, and is more sensitive to just the high density region. For these reasons pions are a very interesting observable. Pions are produced through an intermediate step via production of $\Delta$ resonances. Here the reaction $ NN \leftrightarrow \Delta N$ takes place, where nucleon-nucleon collisions form $\Delta$(1232) baryon resonance from one of the nucleons, which then decays shortly thereafter into a pion and nucelon, $\Delta \leftrightarrow \pi N$. The threshold for  an on-shell $\Delta$ resonance production in nucleon-nucleon collisions, with a mass of \SI{1232}{\mega\electronvolt\per\clight\squared}, corresponds to a laboratory beam of \SI{290}{\MeVA} in a fixed target experiment. In large nuclei, the internal motion of the nucleons is substantial and this modifies the production rates. In the Fermi gas model, nucleons are arranged filling up higher energy levels up until they reach the Fermi energy. The associated Fermi velocity can add to the beam velocity and actually be above the threshold condition, even when the beam velocity is below the threshold \cite{fermiEnergy}.

@@

\section{Motivation}
In an effort to answer the high density behavior of the symmetry energy, we designed a new detector and a set of experiments of Sn + Sn collisions at \SI{270}{\MeVA}. Here we utilized inverse kinematics where the beam is made of a radioactive neutron rich beam, impinging on a stable Sn target. We can probe the isospin asymmetry of the symmetry energy by changing the neutron-proton asymmetry of the incoming beam. Pion production has been studied before in stable beams for beam energies of \SI{400}{\MeVA} and above \cite{fopi}. Yet in these previous experiments, only the total pion yields were published and not the pion spectra. It was also known that the efficiency analysis of the FOPI collaboration \cite{fopi} which was particularly difficult especially for low energy pions. In fact the published pion multiplicities are not the measured values but rather extrapolated to low pion energies. We know now that below the FOPI cutoff cited in \cite{fopi}, approximately 35\% of the $\pi^+$ multiplicity and 55\% of the $\pi^-$ multiplicity, showing the importance of measuring down to low pion energies. While the pion production increases with the beam energy, this increase is countered by the fact that the effects of the symmetry energy decrease with energy \cite{fopi}. This is because available energy to produce pions is larger and therefore less sensitive to the relatively small symmetry energy.  The goal of this Dissertation was to design and build a high efficiency detector in order to measure pion and light charged particle spectra resulting from HICs. To do this a new Time Projection Chamber (TPC) was made called the SAMURAI pion Reconstruction Ion Tracker (\spirit) TPC.

### 2020-08-09 chapters_altFormat/ch_results.tex
Added Betty's edits

Other unconstrained physics includes aspects of the mean field potentials for nucleons, pions, and deltas. A primary focus is on the isovector potentials, of which only nucleon potentials has been varied in the calculations shown below. Some of the models include the momentum dependent nucleonic mean field potentials, and the mean field potentials for deltas and pions. Calculations have already identified that the overall pion yields are influenced by the mean field potentials for the deltas \cite{cozmaPC}, which may not be surprising since a similar sensitivity has govern the occurrence of delta and pionic matter within neutron stars \cite{deltaNS,pionNS}.

@@

We begin with the total charged pion multiplicities for the two systems of extreme isospin asymmetry: the neutron rich $\tin{132}{124}$ system and the neutron deficient  $\tin{108}{112}$ system. Though the previous FOPI data set provided interesting data relevant to  the symmetry energy, they relied on extrapolations to low transverse momenta, $P_T = \sqrt{p_x^2 + p_y^2}$, below \SI{100}{\mega\electronvolt\per\clight} \cite{fopi}. In the \spirit TPC, we were able to measure pions to very low $P_T$ for rapidities $y > y{CM}$; where the rapidity is expressed along the beam direction (z) as,

@@

\begin{equation}
y_o = \frac{y_{\pi CM}}{y_{beam} - y_{CM}},
\end{equation}

@@

The integrated pion yield for both systems, and the $\pi^-/\pi^+$ ratio, is listed in Table~\ref{tb:pionyield}, where the systematic errors are the first error bar and the statistical error is listed next. This represents data with an estimated average impact parameter of $b_{avg} = \SI{2.2}{\femto\metre}$ and $b_{max} < \SI{3}{\femto\metre}$. Notice that the pion ratio is significantly greater than the $N/Z$ of the system.  This is not completely unexpected; in the delta resonance model mentioned in Section~\ref{sec:pionObs}, we expect  $\pi^-/\pi^+ \sim (N/Z)^2$ \cite{baoan_piprod1,baoan_piprod2}, where N/Z is the ratio of neutron to protons in the dense region where pions are produced. For the $\tin{132}{124}$ system $(N/Z)^2$ = 2.56 and for $\tin{108}{112}$ $(N/Z)^2$ = 1.44, but this n\"aive model assumes no migration of nucleons in or out of the high density region where the isovector mean field and other transport properties may a impact that approximation significantly.

@@

The FOPI collaboration has measured the total pion multiplicity resulting from ${}^{197}Au + {}^{197}Au$ collisions at several, higher, beam energies. In the $\tin{132}{124}$ data the asymmetry is $N/Z=1.56$ whereas in the ${}^{197}Au + {}^{197}Au$ the $N/Z=1.493$. Since 4 beams were measured in this experiment, the $N/Z$ dependence was measured as seen in Figure~\ref{fig:totalpiRatio}. Here the dependence is fitted with a 2-nd order polynomial fit. To compare the pion ratio in the $\tin{132}{124}$ data with that of the lower N/Z in the FOPI experiment, we scaled by the pion ratio between $N/Z=1.56$ and 1.493 as given from the fitted polynomial line. Figure~\ref{fig:fopiPionRatio} shows the scaled pion ratio as compared with the FOPI pion ratio \cite{fopi}. The fitted function has the functional form of $p_1(E - p_2)^{-2}$ where $p_1$ and $p_2$ are free parameters, and only is meant to guide the eye. It is also worth mentioning that the pion ratio observed in the FOPI \SI{400}{\MeVA} setting was already considerably higher than what is expected from the $(N/Z)^2$ delta resonance model \cite{baoan_piprod1,baoan_piprod2}. It it worth mentioning that the two experiments differ considerably in their efficiencies as was mentioned before. The FOPI results heavily relied on extrapolations to lower pion energies which the \spirit TPC did not rely on \cite{fopi}. Without a systematic error calculation or discussion with the FOPI collaboration no further detail can be given except that the \spirit TPC data seems to be reasonable expectation and in general agreement with the FOPI data set. Scaling the other systems in the \spirit TPC data leads to almost the exact same scaled value of $\mathrm{Y}(\pi^-)/\mathrm{Y}(\pi^+) = \num{4.82}$ and would be redundant.

>>>>

Other unconstrained physics includes aspects of the mean field potentials for nucleons, pions, and deltas. A primary focus is on the isovector potentials, of which only nucleon potentials has been varied in the calculations shown below. Some of the models include the momentum dependent nucleonic mean field potentials, and the mean field potentials for deltas and pions. Calculations have already identified that the overall pion yields are influenced by the mean field potentials for the deltas \cite{cozmaPC}, which may not be surprising since a similar sensitivity has governed the occurrence of delta and pionic matter within neutron stars \cite{deltaNS,pionNS}.

@@

We begin with the total charged pion multiplicities for the two systems of extreme isospin asymmetry: the neutron rich $\tin{132}{124}$ system and the neutron deficient  $\tin{108}{112}$ system. Though the previous FOPI data set provided interesting data relevant to  the symmetry energy, they relied on extrapolations to low transverse momenta, $P_T = \sqrt{p_x^2 + p_y^2}$, below \SI{100}{\mega\electronvolt\per\clight} \cite{fopi}. In the \spirit TPC, we were able to measure pions to very low $P_T$ for rapidities $y > y_{CM}$; where the rapidity is expressed along the beam direction (z) as,

@@

\begin{equation}
y_o = \frac{y_{\pi_{CM}}}{y_{beam} - y_{CM}},
\end{equation}

@@

The integrated pion yield for both systems, and the $\pi^-/\pi^+$ ratio, is listed in Table~\ref{tb:pionyield}, where the systematic errors are the first error bar and the statistical error is listed next. This represents data with an estimated average impact parameter of $b_{avg} = \SI{2.2}{\femto\metre}$ and $b_{max} < \SI{3}{\femto\metre}$. Notice that the pion ratio is significantly greater than the $N/Z$ of the system.  This is not completely unexpected; in the delta resonance model mentioned in Section~\ref{sec:pionObs}, we expect  $\pi^-/\pi^+ \sim (N/Z)^2$ \cite{baoan_piprod1,baoan_piprod2}, where N/Z is the ratio of neutron to protons in the dense region where pions are produced. For the $\tin{132}{124}$ system $(N/Z)^2$ = 2.56 and for $\tin{108}{112}$ $(N/Z)^2$ = 1.44, but this n\"aive model assumes no migration of nucleons in or out of the high density region where the isovector mean field and other transport properties may impact that approximation significantly.

@@

The FOPI collaboration has measured the total pion multiplicity resulting from ${}^{197}Au + {}^{197}Au$ collisions at several, higher, beam energies. In the $\tin{132}{124}$ data the asymmetry is $N/Z=1.56$ whereas in the ${}^{197}Au + {}^{197}Au$ the $N/Z=1.49$. Since 4 beams were measured in this experiment, the $N/Z$ dependence was measured as seen in Figure~\ref{fig:totalpiRatio}. Here the dependence is fitted with a 2-nd order polynomial fit. To compare the pion ratio in the $\tin{132}{124}$ data with that of the lower N/Z in the FOPI experiment, we scaled by the pion ratio between $N/Z=1.56$ and 1.493 as given from the fitted polynomial line. Figure~\ref{fig:fopiPionRatio} shows the scaled pion ratio as compared with the FOPI pion ratio \cite{fopi}. The fitted function has the functional form of $p_1(E - p_2)^{-2}$ where $p_1$ and $p_2$ are free parameters, and only is meant to guide the eye. It is also worth mentioning that the pion ratio observed in the FOPI \SI{400}{\MeVA} setting was already considerably higher than what is expected from the $(N/Z)^2$ delta resonance model \cite{baoan_piprod1,baoan_piprod2}. It is worth mentioning that the two experiments differ considerably in their efficiencies as was mentioned before. The FOPI results heavily relied on extrapolations to lower pion energies which the \spirit TPC did not rely on \cite{fopi}. Without a systematic error calculation or discussion with the FOPI collaboration no further detail can be given except that the \spirit TPC data seems to be within the expectation and in general agreement with the FOPI data set. Scaling the other systems in the \spirit TPC data leads to almost the exact same scaled value of $\mathrm{Y}(\pi^-)/\mathrm{Y}(\pi^+) = \num{4.82}$ and would be redundant.

### 2020-08-09 chapters_altFormat/ch_summary.tex
Added Betty's edits

  This marks the first time pions have been accurately measured in the sub-threshold region. We observed significant a deviation from the  n\"aive expectation of the simple $\Delta$ resonance model, where $\pi^-/\pi^+ \sim (N/Z)^2$, as discussed in Section~\ref{sec:pionObs}. Such a deviation is also predicted by theory, so this explanation should be retired in favor of something more quantitative and informative. Obtaining a better explanation from current calculations is difficult because nearly all pion production calculations are incomplete and each model has a mixture of strengths and inadequacies that are unique to each model. Thus, it is difficult to get the theorists to agree about the consequences of some of the model assumptions that they employ; and consequently a constraint on the density dependence of the symmetry energy.

>>>>

  This marks the first time pions have been accurately measured in the sub-threshold region. We observed a significant deviation from the  n\"aive expectation of the simple $\Delta$ resonance model, where $\pi^-/\pi^+ \sim (N/Z)^2$, as discussed in Section~\ref{sec:pionObs}. Such a deviation is also predicted by theory, so this explanation should be retired in favor of something more quantitative and informative. Obtaining a better explanation from current calculations is difficult because nearly all pion production calculations are incomplete and each model has a mixture of strengths and inadequacies that are unique to each model. Thus, it is difficult to get the theorists to agree about the consequences of some of the model assumptions that they employ; and consequently a constraint on the density dependence of the symmetry energy.
